---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Loading needed libraries and initializing the environment

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
library(knitr)
library(dplyr)
library(ggplot2)
```


## Loading and preprocessing the data

```{r Load & read csv file, }
setwd("/Users/costal3/projects/R/data_analysis/reproducible-research/Course_project_1")
raw_data<-read.csv("activity.csv")
head(raw_data)
```
Remove all NA's, add calculated day, calculate summarised steps/day

```{r}
activity<-na.omit(raw_data)
activity$data <- as.Date(activity$date)
steps_per_day <- summarise(group_by(activity, date), total = sum(steps))

hist(steps_per_day$total,
   main="Histogram of total number of steps per day",
   xlab="Total # of steps / day")
mean_steps<-mean(steps_per_day$total)
median_steps <- median(steps_per_day$total)
```


## What is mean total number of steps taken per day?

- Mean total number steps per day is `r mean_steps`  
- Median total number of steps per day is `r median_steps`  
- Data recorded between `r min(as.Date(activity$date))` and `r max(as.Date(activity$date))`  

## What is the average daily activity pattern?

``` {r}
average_steps_per_day <- aggregate(steps~interval, activity, FUN="mean")
plot(average_steps_per_day,
   type="l",
   main="Average number of steps over whole time period",
   ylab="Average steps taken")
max_step_interval <- average_steps_per_day$interval[which.max(average_steps_per_day$steps)]
```

- The highest average step count happened during interval `r max_step_interval`

## Imputing missing values
- Calculate the number of missing observations
- Fill in missing data with medians
```{r}
total_missing_values <- sum(is.na(raw_data$steps))
```
- There are `r total_missing_values` missing observations (NAs) in the data
- For those NA's, let's calculate the observed average steps from similar intervals
``` {r}
edited_data <- raw_data
for (i in 1:nrow(edited_data)) {
   if (is.na(edited_data$steps[i])) {
      #Find average steps for this interval (i) and provision on edited_data$steps
      j <- which (edited_data$interval[i] == average_steps_per_day$interval)
      edited_data$steps[i] <- average_steps_per_day[j, ]$steps
   }
}
```
- Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day
```{r}
steps_per_day <- summarise(group_by(edited_data, date), total = sum(steps))

hist(steps_per_day$total,
     main="Histogram of total number of steps per day (edited data)",
     xlab="Total # of steps / day")
mean_steps <- mean(steps_per_day$total)
median_steps <- median(steps_per_day$total)
```
- Mean total number steps per day is `r mean_steps` (same as before)  
- Median total number of steps per day is `r median_steps` (slighlty higher)  

## Are there differences in activity patterns between weekdays and weekends?
```{r}
#Mark weekday/weekend using function weekdays and provision on day_type
edited_data$day <- weekdays(as.Date(edited_data$date))
edited_data$day_type[edited_data$day %in% c("Saturday", "Sunday")] <- "Weekend"
edited_data$day_type[is.na(edited_data$day_type)] <- "Weekday"

#calculate new average steps per day
plot_data<-aggregate(edited_data$steps, by=edited_data[c("day_type", "interval")], FUN="mean")
names(plot_data)<-c("day_type", "interval", "Average_steps")

#Plot it
qplot(interval, 
   Average_steps, 
   data = plot_data, 
   geom=c("line"),
   xlab = "Interval", 
   ylab = "# of steps", 
   main = "") +
   facet_wrap(~ day_type, ncol = 1)
```

